Make any stupid plot you want

### 2021-01-21 draft/sections/math.tex
reshuffling data/moving around and kicking out continuity:

We use fiber bundles\cite{FiberBundle2020, rowlandFiberBundle} to model data and graphics because they allow us to seperate the variables in a dataset from how the values  are connected to each other \cite{butlerVisualizationModelBased1989,butlerVectorBundleClassesForm1992}. The fiber bundles mentioned in this work are assumed to be locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre} unless otherwise specified.

@@

\subsection{Data Space}
\label{sec:data}
Lets say we are working with the Global Historical Climatology Network\cite{AnOverviewoftheGlobalHistoricalClimatologyNetworkDailyDatabase}, which is a dataset of global daily meterological observations. Each record includes the timestamp, location, and measurements such as temperature, precipitation, and snowfall. Using a fiber bundle allows us to express the properties of the variables(\ref{sec:data_fiber}), the connectivity of the records (\ref{sec:data_base}), and tables of records (\ref{sec:data_section}).

\subsubsection{Variables: Fiber Space $F$}
\label{sec:data_fiber}
The fiber is a topological space containing the set of all records that could possibly be in the dataset, even values that do not actually exist in the dataset. For example, a fiber $F$ that is the temperature variable from GHCN is the range of physically possible temperature values $ U_{temp} = \{x \in \mathbb{R}^+\}$. 
\begin{equation}
    F = U_{temp}
\end{equation}
Then lets add the time variable from GHCN $U_{time} = \{x \in \mathbb({Z}^+ \mid x \geq 1011832\}$ 
\begin{equation}
F =  U_{temp} \times U_{time}
\end{equation}
such that $F$ is the cartesian product of $F_0=U_{temp}$ and $F_1 = U_{time}$. The elements in a given fiber set do not need to be scalers; for example the location of each record can be encoded as $U_{loc} = \{y, x \in \mathbb{R}^{2} \mid -90 \leq y \leq 90, -180 \leq x -180\}$: 
\begin{equation}
    F = U_{temp} \times U_{time} \times U_{loc}
\end{equation}
such that one record $r$ in the fiber would have the form $r=(temperature,\, time,\, (latitude,\, longitude))$. We also add the variable name and measurement scale $M$ monoid actions to the fiber $F_i$ such that 
\begin{align}
    F_0 &= (\textit{temperature},\, U_{temp},\, M_0)\\
    F_1 &= (\textit{time},\, U_{time},\, M_1)\\
    F_2 &= (\textit{location},\, U_{loc},\, M_2)
\end{align}
 or more generally (name, type, monoid action)
 \begin{equation}
    F_i = (\textit{name},\, U_{type},\, M_{i})
 \end{equation}
where type is the set of values defined by the datatype as described in Spivak's categorical formulation of databases \cite{spivakSIMPLICIALDATABASES}. We describe $M$ in more detail in section~\ref{sec:data_monoid}. We can keep growing the fiber
\begin{equation}
F= F_{0} \times \ldots \times F_{i}\times\ldots\times F_{n}
\end{equation}
and identify the elements in the record by the variable name. We will often be working with the decoupled fiber $F_i$ rather than the total fiber $F$.

@@

\begin{figure}[ht!]
    \includegraphics[width=\textwidth]{figures/math/rotation_actions.png}
    \label{fig:data_monoid_rotation}
    \caption{The set of rotation actions [turn 0, 90, 180] on the star produce a star (closed), include a rotation that does not change the orientation (identity), and can be done in any order to produce the same final orientation (closure); therefore the rotations are a set of monoid actions.}.  
\end{figure}

A common task is modifying the values in $F_i$ in some consistent way, such as converting data from \textdegree F to \textdegree C or changing category labels or rotating a shape such as in figure~\ref{fig:data_monoid_rotation}. This model formally encodes the set of allowable operations in the fiber because a transformation on the data side needs to be reflected in the visual representation, discussed in more detail in section~\ref{sec:artist_nu}. One method of classifying the types of operations that can be performed on data are Steven's measurment scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale}; nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.

We generalize the set of allowable transformations to the monoid actions $M_i$ on $F_i$. A monoid \cite{Monoid2021} $M_i$ is a set with an associative binary operator $\ast:M_i \times M_i\rightarrow M_i$. A monoid has an identity element $e\in M_i$ such that $e\ast a= a \ast e = a$ for all $a \in M_i$. A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M_i$ is a set $U_{type}$ with an action $\bullet: M\times U_{type} \rightarrow U_{type}$ with the properties:
\begin{align*}
    \textbf{associativity}\;& \text{for all } m,t \in M \text{ and } x\in U_{type}, m\bullet(t\bullet x) = (m\ast t) \bullet x\\
    \textbf{identity}\;& \text{for all } x\in U_{type}, e\in M,  e\bullet x = x 
\end{align*}

For example, the temperature $F_0$ defined in unit \textdegree C is on an interval scale; this means that the monoid actions $M_0$ on $U_{temp}$ are the set of transformations $\{x + c \in U_{temp} \mid x,c \in U_temp\}$ where $c$ is a valid interval between two temperatures $c=x_i-x_j$.

As with the fiber $F$, the total monoid space $M$ is the cartesian product
\begin{equation}
M = M_{0} \times \ldots \times M_{i}\times \ldots \times\ldots M_{n}
\end{equation}
of the monoid $M_{i}$ on each $F_{i}\in F$. Because each $F_i, M_i$ can be independent of the other fibers, we can formally express the constraints on heterogenous multivariate datasets in the same manner as univariate datasets.

\subsubsection{Connectivity: Base Space $K$} 
\label{sec:data_base}
The fiber provides the schema of the dataset, but the fiber bundle is what glues that schema to a structure. The fiber bundle is the map $\pi$ between the fiber space included in the total space $E$ and the base space $K$
\begin{equation}
    \label{eq:fiber_bundle}
    \begin{tikzcd}
        F \arrow[r, hook] & E \arrow[r, "\pi"] & K
    \end{tikzcd}
\end{equation}

@@

\begin{figure}
    \label{fig:data_base_transition}
    \includegraphics[width=\textwidth]{figures/math/transition_functions.png}
    \caption{Many non-trivial spaces can be made locally trivial by dividing $E$ into locally trivial subspaces and defining transition functions between the edges on K for how to glue the two subspaces such that the sigmas are continuous.}
\end{figure}

A fiber bundle is trivial if $E = K\times F$, which means that the fibers $F$ are the same over every $k$ \cite{rowlandFiberBundle,FiberBundle2020}. We can make many non-trivial spaces locally trivial by breaking up the total space $E$ over local neighborhoods $U \subset K$ such that the fibers are equivalant. An example of a non-trivial fiber bundle is a mobius strip, where because of the twist some fibers are $\left[0,1\right]$ and others are $\left[1,0\right]$ such that we cannot construct a global map of which edge is 0 and which edge is 1.  In figure ~\ref{fig:data_base_transition}, the mobius strip is cut into two rectangular base spaces over $K=\{e_1, e_2\}$. We define transition functions from the fiber to the fiber along the edges of the rectangles such that we can glue these rectangles together to reform the mobius strip. A constraint we impose on the transition functions is that the monoid actions are commutative
-% need more detail please on commutativty

@@

The sections of the fiber bundle are maps from points $k$ on $K$ to records $r$ in $F$. When we map our data to visual variables, the $\tau$ returns the set of values that are mapped together into a single graphic element. For GHCN, the $\tau$ functions return a table in a database with the schema defined in $F$ and the connectivity defined in $K$: this means the sections can be defined such that all values have the same year or location, e.g each table returned by a $\tau$ is for a different station.

The section $\tau$ is the mapping from base space to total space $\tau: K\rightarrow E$ 
\begin{equation}
    \label{eq:section}
    \begin{tikzcd}
        F \arrow[r, hook] & E \arrow[d, "\pi"']           \\
        & K \arrow[u, "\tau"', bend right]
    \end{tikzcd}
\end{equation}
such that there is $\pi(\tau(k)) = k$ map back to $K$.

If the fiber bundle is locally trivial, then
\begin{equation}
    \label{eq:section_return}
    \tau(k) = (k, (g_{F_{0}}(k), \ldots, g_{F_{n}}(k))
\end{equation}
which is also true when the fiber bundle is locally trivial over $U \subset K$.
%%%can decompose bundles into Ei such that \tau: E_{i} \rightarrow K, where F_{i} is the fiber of E_{i}
which we can also decompose such that
\begin{align}
\tau_{0}(k) &= (k, (x_{F_{0}}))\\
            & \vdotswithin{=}\\
\tau_{n}(k) &= (k, (x_{F_{n}}))
\end{align}

which allows us to have field wise access to the data so long as there is a shared $K$.  
%% we will use this property later for visual mapping such that we can seperate each variable into its own bundle

@@

% can't make a 1D fiber - can't make a 1D fiber, switch the picture to plane over 0D and 1D and then can talk about disconnected or connected

@@

\subparagraph{Example}
\begin{figure}[ht!]
    \label{fig:triangle_torus}
    \includegraphics[width=\textwidth]{figures/math/triangle_torus.png}
    \caption{Representation of the base space $K$ of a toridal total space $E$ as two connected 2-simplices.}
\end{figure}

>>>>

@@

\subsection{Data Space}
\label{sec:data}
%%% maybe change this to a generic representational data set
We build on Butler's fiber bundle driven abstraction of data formats\cite{butlerVectorBundleClassesForm1992,butlerVisualizationModelBased1989} because fiber bundles can express the properties of the variables(\ref{sec:data_fiber}), the continuity of the records (\ref{sec:data_base}), and distinct instances of records (\ref{sec:data_section}). The fiber bundles mentioned in this work are assumed to be locally trivial\cite{spanier1989algebraic,LocallyTrivialFibre} unless otherwise specified.

\subsubsection{Variables: Fiber Space $F$}
\label{sec:data_fiber}
The fiber is a topological space containing the set of all records that could possibly be in the dataset, even values that do not actually exist in the dataset.

%%push straight math generalization up
%% We define w/ mapping, as Spivak

%%push down to an example, knock it back to semi concrete
For example, a fiber $F$ that is the temperature variable from GHCN is the range of physically possible temperature values $ U_{temp} = \{x \in \mathbb{R}^+\}$. 
\begin{equation}
    F = U_{temp}
\end{equation}
Then lets add the time variable from GHCN $U_{time} = \{x \in \mathbb{R}^+\}$ 
\begin{equation}
F =  U_{temp} \times U_{time}
\end{equation}
such that $F$ is the cartesian product of $F_0=U_{temp}$ and $F_1 = U_{time}$. The elements in a given fiber set do not need to be scalers; for example the location of each record can be encoded as $U_{loc} = \{y, x \in \mathbb{R}^{2} \mid -90^\circ \leq y \leq 90^\circ, -180^\circ \leq x -180^\circ\}$: %add comment about latitude and longitude
\begin{equation}
    F = U_{temp} \times U_{time} \times U_{loc}
\end{equation}
such that one record $r$ in the fiber would have the form $r=(temperature,\, time,\, (latitude,\, longitude))$. We also add the variable name and measurement scale $M$ monoid actions to the fiber $F_i$ such that 
\begin{align}
    F_0 &= (\textit{temperature},\, U_{temp},\, M_0)\\
    F_1 &= (\textit{time},\, U_{time},\, M_1)\\
    F_2 &= (\textit{location},\, U_{loc},\, M_2)
\end{align}%the fiber is the set, variable is the tuple such that there is a map from the index i in F to the tuple (name, M_0)  for every U_{type} we can construct a map to the name and the type as described by Spivak 
 or more generally (name, type, monoid action)
 \begin{equation}
    F_i = (\textit{name},\, U_{type},\, M_{i})
 \end{equation}
where type is the set of values defined by the datatype as described in Spivak's categorical formulation of databases \cite{spivakSIMPLICIALDATABASES}. We describe $M$ in more detail in section~\ref{sec:data_monoid}. We can keep growing the fiber
\begin{equation}
F= F_{0} \times \ldots \times F_{i}\times\ldots\times F_{n}
\end{equation}
and identify the elements in the record by the variable name. We will often be working with the decoupled fiber $F_i$ rather than the total fiber $F$.

%%invert this section, start w/ general math than GHCN

@@

\begin{figure}[ht!]
    \includegraphics[width=\textwidth]{figures/math/rotation_actions.png}
    \label{fig:data_monoid_rotation}
    \caption{The set of rotation actions [turn 0, 90, 180] on the star produce a star (closed), include a rotation that does not change the orientation (identity), and can be done in any order to produce the same final orientation (closure); therefore the rotations are a set of monoid actions.
    %% dataset that is symmetric, fiber is x,y, rotatation action
    %% rotate + plot or plot + rotate
    %%emoji shuffle 
    }  
\end{figure}

A common task is modifying the values in $F_i$ in some consistent way, such as converting data from \textdegree F to \textdegree C or changing category labels or rotating a shape such as in figure~\ref{fig:data_monoid_rotation}. This model formally encodes the set of allowable operations in the fiber because a transformation on the data side needs to be reflected in the visual representation, discussed in more detail in section~\ref{sec:artist_nu}.

We generalize the set of allowable transformations to the monoid actions $M_i$ on $F_i$. A monoid \cite{Monoid2021} $M_i$ is a set with an associative binary operator $\ast:M_i \times M_i\rightarrow M_i$. A monoid has an identity element $e\in M_i$ such that $e\ast a= a \ast e = a$ for all $a \in M_i$. A left monoid action \cite{SemigroupAction2021,ActionNLab} of $M_i$ is a set $U_{type}$ with an action $\bullet: M\times U_{type} \rightarrow U_{type}$ with the properties:
\begin{align*}
    \textbf{associativity}\;& \text{for all } m,t \in M \text{ and } x\in U_{type}, m\bullet(t\bullet x) = (m\ast t) \bullet x\\
    \textbf{identity}\;& \text{for all } x\in U_{type}, e\in M,  e\bullet x = x 
\end{align*}
As with the fiber $F$, the total monoid space $M$ is the cartesian product
\begin{equation}
M = M_{0} \times \ldots \times M_{i}\times \ldots \times\ldots M_{n}
\end{equation}
of the monoid $M_{i}$ on each $F_{i}\in F$.

Steven's defined the measurment scales \cite{stevensTheoryScalesMeasurement1946,leaFormalizationMeasurementScale} in terms of the monoid actions permissible on the measurments; nominal data is permutable, ordinal data is monotonic, interval data is translatable, and ratio data is scalable \cite{weissteinSimilarityTransformation}.

For example, the temperature in \textdegree C is on an interval scale, which means that the monoid actions $M_0$ on $U_{temp}$ are the set of translations $\{f: x\mapsto x + c \in U_{temp} \mid x,c \in U_temp\}$ where $c$ is a valid interval between two temperatures $c=x_i-x_j$.

\subsubsection{continuity: Base Space $K$} 
%%fold butlder
\label{sec:data_base}
The fiber provides the schema of the dataset, but the fiber bundle is what glues that schema to a structure. The fiber bundle is the map $\pi$ between the fiber space included in the total space $E$ and the base space $K$
\begin{equation}
    \label{eq:fiber_bundle}
    \begin{tikzcd}
        F \arrow[r, hook] & E \arrow[r, "\pi"] & K
    \end{tikzcd}
\end{equation}

@@

@@

The sections of the fiber bundle are maps from points $k$ on $K$ to records $r$ in $F$. When we map our data to visual variables, the $\tau$ returns the set of values that are mapped together into a single graphic element. The section $\tau$ is the mapping from base space to total space $\tau: K\rightarrow E$ 
\begin{equation}
    \label{eq:section}
    \begin{tikzcd}
        F \arrow[r, hook] & E \arrow[d, "\pi"']           \\
        & K \arrow[u, "\tau"', bend right]
    \end{tikzcd}
\end{equation}
such that there is $\pi(\tau(k)) = k$ map back to $K$.

A fiber bundle is trivial if $E = K\times F$\cite{rowlandFiberBundle,FiberBundle2020}.
All fiber bundles are locally trivial, which means if you restrict $E$ over a small enough neighborhood $U \subset K$ then it is a locally trivial bundle over $U$.
When the fiber bundle is locally trivial, then
\begin{equation}
    \label{eq:section_return}
    \tau(k) = (k, (g_{F_{0}}(k), \ldots, g_{F_{n}}(k))
\end{equation}
the section $\tau$ is the record in the $F$. Because we can decompose the bundle and the fiber, we can formulate $\tau$ as 
\begin{align}
\tau = (\tau_0,\ldots, \tau_i, \dots, \tau_n) 
\end{align}
where each section $\tau_i$ is a variable or set of variables.

\paragraph{Example?}
For GHCN, the $\tau$ functions return a table in a database with the schema defined in $F$ and the continuity defined in $K$: this means the sections can be defined such that all values have the same year or location, e.g each table returned by a $\tau$ is for a different station.

@@

Lets say we are working with the Global Historical Climatology Network\cite{AnOverviewoftheGlobalHistoricalClimatologyNetworkDailyDatabase}, which is a dataset of global daily meterological observations. Each record includes the timestamp, location, and measurements such as temperature, precipitation, and snowfall.
% can't make a 1D fiber - can't make a 1D fiber, switch the picture to plane over 0D and 1D and then can talk about disconnected or connected

@@

\subparagraph{Example: Mobius Strip}\mbox{}\\
\begin{figure}[ht!]
    \label{fig:data_base_transition}
    \includegraphics[width=\textwidth]{figures/math/transition_functions.png}
    \caption{Many non-trivial spaces can be made locally trivial by dividing $E$ into locally trivial subspaces and defining transition functions between the edges on K for how to glue the two subspaces such that the sigmas are continuous.}
\end{figure}

 An example of a non-trivial fiber bundle is a mobius strip\cite{find the citation for why};

In figure ~\ref{fig:data_base_transition}, the mobius strip is cut into two rectangular base spaces over $K=\{e_1, e_2\}$. We define transition functions from the fiber to the fiber along the edges of the rectangles such that we can glue these rectangles together to reform the mobius strip. A constraint we impose on the transition functions is that the monoid actions are commutative
%monoid action m*f\mapsto delta(mf)= m*delta(f)

\subparagraph{Example: Torus}\mbox{}\\
\begin{figure}[ht!]
    \label{fig:triangle_torus}
    \includegraphics[width=\textwidth]{figures/math/triangle_torus.png}
    \caption{Representation of the base space $K$ of a toridal total space $E$ as two connected 2-simplices.}
\end{figure}
